---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Preparation of environment
```{r preparation_environment, message = FALSE}
library(dplyr)
library(lattice)
library(lubridate)
library(RColorBrewer)
pal1 <- brewer.pal(9, "Pastel1")
pal2 <- brewer.pal(9, "Set1")
```

## Loading and preprocessing the data

```{r loading_preprocessing_data}
options(scipen = 1, digits = 2)
data <- read.csv("activity.csv")
data1 <- data %>%
	group_by(date)	%>%
	summarise(totalSteps = sum(steps))
hist(data1$totalSteps, main = "Total number of steps per day", 
	 xlab = "Number of steps per day", col = pal1[2])
```


## What is mean total number of steps taken per day?

```{r total_steps_day}
meanSteps   <- mean(data1$totalSteps, na.rm = TRUE)
medianSteps <- median(data1$totalSteps, na.rm = TRUE)
```

The mean total number of steps taken per day is **`r meanSteps`**.

The median total number of steps taken per day is **`r medianSteps`**.

## What is the average daily activity pattern?

```{r daily_activity_pattern}
data2 <- data %>%
	group_by(interval) %>%
	summarise(meanSteps = mean(steps, na.rm = TRUE))
with(data2, plot(interval, meanSteps, main = "Average daily activity pattern",
	xlab = "5-minutes time interval", ylab = "Average number of steps per day",
	col = pal2[2], type = "l", lwd = 2))
maxInterval <- data2[which.max(data2$meanSteps),1][[1]]
```

The **`r maxInterval`** 5-minute interval contains the maximum number of steps, on average across all the days.

## Imputing missing values

```{r imputing_missing_values}
missingValues <- sum(!complete.cases(data))
# Extrapolate missing values from mean intervals
extrapolatedInterval <- lapply(data$interval, function(x) {
	data2[which(data2$interval == x),2][[1]] }) %>%
	matrix(ncol = 1, dimnames = list(NULL, "extrapolatedInterval")) %>%
	data.frame(stringsAsFactors = FALSE) 
# Original data with missing values replaced by mean intervals
data <- data %>%
	bind_cols(extrapolatedInterval) %>%
	mutate(steps = ifelse(
		is.na(steps), as.numeric(extrapolatedInterval), steps))	
data1 <- data %>%
	group_by(date) %>%
	summarise(totalSteps = sum(steps))
hist(data1$totalSteps, 
	 main = "Total number of steps per day - No missing values", 
	 xlab = "Number of steps per day", col = pal1[2])
meanSteps   <- mean(data1$totalSteps, na.rm = TRUE)
medianSteps <- median(data1$totalSteps, na.rm = TRUE)
```

There are **`r missingValues`** missing values in the data set.

The mean total number of steps taken per day is **`r meanSteps`**.

The median total number of steps taken per day is **`r medianSteps`**.

## Are there differences in activity patterns between weekdays and weekends?

```{r weekday_pattern}
data <- data %>%
	mutate(day = ifelse(
		wday(ymd(date)) == 7 | wday(ymd(date)) == 1, "weekend", "weekday")) %>%
	mutate(day = factor(day))
data1 <- data %>%
	group_by(day, interval) %>%
	summarise(averageSteps = mean(steps))
xyplot(averageSteps ~ interval | day, data1, type = "l", layout = c(1,2),
	   xlab = "Interval", ylab = "Number of steps", 
	   col = pal2[[2]], lwd = 2)

```
